---
title: "Functions and Loops  with Vectorizations and Functionals"
author: "Jasmine Kobayashi"
date: "2022-08-29"
output: html_document
---

A related function that is useful is `ifelse`. This function takes in 3 arguments: a logical and two possible answers.

If the logical is 'true', then the value in the **second** argument is returned.

If the logical is `false`, then the value in the **third** argument is returned.

```{r}

a <- 0
ifelse(a>0, 1/a,NA)

```

`ifelse` is useful especially when working with vectors. It examines each entry of the logical vector and returns elements from the vector provided in the second argument, if the entry is `true`, or elements in from the vector in the third argument if the entry is `false`. 


```{r}

a <- c(0, 1, 2, -4, 5)
result <- ifelse(a>0, 1/a, NA)
result

```


```{r}
library(dslabs)
data(na_example)

no_nas <- ifelse(is.na(na_example), 0, na_example)
sum(is.na(no_nas))

```

Two other useful functions are `any` and `all`. The `any` function takes a vector of logicals and returns `TRUE`. The `all` function takes a vector of logicals and returns `TRUE` if all the entries are `TRUE`.

```{r}

z <- c(TRUE,TRUE, FALSE)
any(z)
all(z)
```

# Defining Functions (read book chapter)

```{r}

avg <- function(x){

s <-  sum(x)
n <- length(x)
s/n
}

```

Note that `avg` is a function that computes the mean.

```{r}

x <- 1:100
identical(mean(x), avg(x))

```

# For Loops

Partial summation Example $S_n$

```{r}

compute_s_n <- function(n){
  
  x <- 1:n
  sum(x)
  
}

```


How can we compute $S_n$ for various values of $n$, such as $n=1,...,15$?.

```{r}

for(i in 1:25)
  print(i)

```


Formula for partial sum of a series is given by $$\frac{n(n+1)}{2}$$

For loops, let us define the range that our variable takes, then change the value and evaluate the expression as we run our loop. 


```{r}
for(i in 1:10)
  print(i)
```

Here is the for-loop we could write for our $S_n$ example:
```{r}

m <- 25
s_n <- vector(length = m)    # creating an empty vector
for (n in 1:m){
  
  s_n[n] <- compute_s_n(n)
  
}

```

In each iteration $n=1 , n=2, etc$ we compute $S_n$ and store it in the *nth* entry of $S_n$. Now we can create a plot to search for a pattern.

```{r}

n <- 1:m
plot(n,s_n)

```

# Vectorization and Functionals

Vectorization over loops shortens and clears up your code. A vectorized function is a function that will apply the same operation on each vectors.

```{r}

x <- 1:10
sqrt(x)

y <- 1:10
x*y

```

Not all functions work this way. For example, let's consider `comupte_s_n`: This function does not work element wise, because it is expecting a scalar. This piece of code does not run the function on each entry of `n`. 

```{r}

n <- 1:25
compute_s_n(n)    #outputs a warning that only first element is calculated

```

```
Functionals are functions that help us apply the same function to each entry in a vector, matrix, data frame or a list. Here, we cover the function that operates on a numeric, logical and character vectors

```

THe function that will utilize that can accomplish this is called `sapply` ("s-apply").  This function permits us to perform element-wise operations on any function. 

```{r}

x <- 1:10
sapply(x,sqrt)

```

Each element of `x` is passed on to the function `sqrt` and the result is returned. These results are concatenated. In this case, the result is a vector of the same length as the original `x`. Therefore, the for-loop above can be written as: 

```{r}

n <- 1:25
s_n <- sapply(n,compute_s_n)

```

(Other functionals:)`apply`, `lapply`, `tapply`, `mapply`, `vapply`, and `replicate`. The main ones we are going to use are `sapply`, `apply`, and `replicate`. 

