---
title: "Introduction to R-Markdown"
author: "Jasmine Kobayashi"
date: "08/28/2023"
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
```
# Data Wrangling and Tidy Data

We will focus on data wrangling, including importing data, changing data types, sub-setting variables and observations, and calculating basic functions. 

## Question 1: Importing Data
Download the California foreclosure auctions dataset and save it to a folder on your desktop. Then load the data into R and save it as a data frame.
```{r Q1-read_data}
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))  # set working directory to current file directory
data <- read.csv("California Foreclosure Auctions_updated.csv")
```

## Question 2: Check out the Data
What is the observational unit of this data? Is the data in tidy format? What needs to be cleaned up about this data?
```{r Q2-Check_data}
(head(data))
(colnames(data))
(summary(data))
```
**JK answer:**
From what I understand about what's considered "tidy", I don't believe this is "tidy". What I understand is what's considered "tidy" in R (as opposed to "clean data" in other senses), the data has rows for each entry of data and columns for all potential variables; in that sense, this data qualifies for all of those. But need to also have entries in all cells and there are definitely some missing data.

## Question 3: Subset Rows and Columns
Create a new data frame that contains only the auctions that are "TrusteesDeed-3rd" type, and only the variables required to uniquely identify each observation, along with "TransferAmount". (Bonus: make a scatter plot of the estimated value vs the final sale price and color code it by County - make some interpretations)

```{r Q3-subset}
# new dataframe of TransferType
# (I'm guessing "uniquely identify each observation" is implying the index of the data)
td3 <- data %>% filter(TransferType=="TrusteesDeed-3rd") %>% select(TransferType,TransferAmount)
head(td3)

```

```{r Bonus-scatter_plot,warning=FALSE,message=FALSE}
# Bonus scatter plot (Estimated value vs Final Price; color coded by county)

# parse_number() (from readr) apparently converts to numeric while ignoring "$"s & commas
data$TransferAmount <- parse_number(data$TransferAmount)  
data$EstimatedValue <- parse_number(data$EstimatedValue)

# scatter plot
data %>% filter(TransferType=="TrusteesDeed-3rd") %>% 
          ggplot(aes(x=TransferAmount,y=EstimatedValue,col=County)) + 
          geom_point() 
```

## Question 4: Tidy the Data
If we wish to treat the County as the observational unit, then we must tidy the data. We can do this by averaging across all observations within the new observational unit. Create a new data.frame that contains the County variable along with average transfer price.

```{r Q4-Tidy}
tidy_county <- data %>% 
                group_by(County) %>% 
                mutate(avg_transfer = mean(TransferAmount,na.rm = TRUE)) %>%   
                select(County,avg_transfer)

head(tidy_county)
```

## Question 5: Calculate By Groups
What is the average sale price for "TrusteesDeed-3rd" transfers for each county? Which county is the most expensive?
```{r Q5-calculate_groups}
tidy_county_td3 <- data %>% 
                  filter(TransferType=="TrusteesDeed-3rd") %>% 
                  group_by(County) %>% 
                  mutate(avg_transfer = mean(TransferAmount,na.rm = TRUE)) %>% 
                  select(County,avg_transfer)

distinct(tidy_county_td3)
```


## Final Questions:

1) What was the main point of today's content?

To review relatively basic data reading, EDA, and tidying in R. 

2) What is the muddiest point of the class so far?

I use more Python than R, so I often need to remind myself of many coding syntax in R. 
(But fortunately I love linear algebra so at least I don't have to worry about that aspect as much! :D)

### Knit the File to HTML and Submit!! 
