---
title: "Analysis of Variance (ANOVA)"
---
```{r libraries, warning=FALSE, message = FALSE}
#Load Tidyverse
library(tidyverse)

#load dplyr
library(dplyr)

#load ggplot2
library(ggplot2)
```

Today we discuss hypothesis testing about samples from about more than two populations.

This analysis is called Analysis of Variance (ANOVA).

Analysis of variance (ANOVA) is a method of comparing means (behavior) across several populations using several samples based on variations from the mean.

If variations from the mean are different for the populations it is unlikely that the population means are the same.

We use more than two samples when we have more than two populations.

We will discuss the following cases:

- **Case 1:** testing hypothesis when samples are independent of each other
- **Case 2:** testing hypothesis when samples are dependent on each other

# Case 1. testing hypothesis when samples are independent of each other

## Example 1:
Suppose we have the following weight samples from three different populations.
The samples drawn from these populations are likely independent since they are drawn from three different populations.
```{r}
smpl1 <-c(95,91,89,90,99,88,96,98,95)

smpl2 <-c(83,89,85,89,81,89,90,82,84,80)

smpl3 <-c(68,75,79,74,75,81,73,77)
```

Suppose we want to check if the population means are likely the same or whether they are they different?

We can test the following null hypothesis at 5% significance level.

- $H_0$: population mean 1 = population mean 2 = population mean 3
- $H_1$: Not $H_0$

- Set $\alpha = 5$% (0.05)

The procedure for testing more than two populations using more than two samples is called Analysis of Variance (ANOVA).

The R function for ANOVA is `aov()`

The syntax for `aov()` is as follows:
```
aov(Dependent_variable ~ Independent_variable, data)
```

Suppose we have drawn three samples of weights from three populations:
```{r}
smpl1 <-c(95,91,89,90,99,88,96,98,95)

smpl2 <-c(83,89,85,89,81,89,90,82,84,80)

smpl3 <-c(68,75,79,74,75,81,73,77)
```

Our goal is to answer the question: are weights on average the same or different across the three populations from where we obtained the samples if we assume equal variances across the three populations?

We have to convert the data into a data frame to use `aov()`

### Step 1: create a single vector that consists of smpl1,smpl2,smpl3
```{r}
single_vector <-c(smpl1,smpl2,smpl3)

single_vector #Note all numbers are now in one variable single_vector
```

### Step 2: create a vector consisting of the sample names matches up against the respective numbers

i.e. a vector consisting of "smpl1" repeated 9 times, "smpl2" repeated 10 times and "smpl3" repeated 8 times.

```{r}
long_vector <-rep(c("smpl1", "smpl2", "smpl3"),
                  times=c(length(smpl1),length(smpl2), length(smpl3)))

long_vector #entries created to help identify the numbers in single_vector
```

### Step 3: create data frame:
```{r}
sample_df <-data.frame(long_vector,single_vector)
```

### Step 4: view the data frame, `sample_df`
```{r}
sample_df
```

Note that the sample now has the structure of a data frame

### Step 5: conduct analysis of variance 
```{r}
anova_analysis <-aov(single_vector ~ long_vector, data=sample_df)
```

### Step 6: Use summary function to display ANOVA results
```{r}
summary(anova_analysis)
```

***Write your finding/result:***

Compare p-value with your chosen alpha

We see that the p-value is $\le \alpha$ (5%) - so we reject the null ($H_0$)

***Write your conclusion:***

We conclude that there is evidence in favor of the alternative hypothesis 

i.e. Not $H_0$

i.e the means of the populations (from where the samples were drawn) are not the same

i.e. the average behavior of the populations are not the same


The R function `oneway.test()` also lets us achieve the same results from `aov()` above

The test above is also called one way ANOVA when we assume the variance from the means for all populations are equal.
```{r}
oneway.test(single_vector ~ long_vector, data=sample_df, var.equal=TRUE)
```

The advantage of the `oneway.test()` function is that you write fewer code 
(e.g you do not need a summary statement to display your ANOVA summary)

## Example 2: Energy drink consumption levels and energy activity levels

Imagine we wanted to test the effects of energy drinks on students in a university population.
So we ask a sample of students to consume energy drinks in various quantities.
We obtain the following data that captures consumption category and energy level after consumption.

Our goal is to answer the question: are energy levels the same across energy drink consumption levels if we assume equal variances for the populations?

```{r}
energy_level <-c(3,2,1,1,4,5,2,4,2,3,7,4,5,3,6)
consumption_level<-gl(3,5, labels = c("No consumption", "Low consumption", "High consumption"))

study_data <-data.frame(consumption_level,energy_level)
```

The `gl()` function is used to generate factor levels - essentially

create categorical/level values for the categorical variable

One syntax of the `gl()` function:
```
gl(n, k, length = n*k, labels = seq_len(n), ordered = FALSE)
```

Where:

- `n`: This is an integer that indicates the number of levels
- `k`: This is an integer that indicates the number of replications
- `length`: This is an integer that indicates the length of the result
- `labels`: This is an optional vector of labels for the output of the factor levels
- `ordered`: This takes a logical value (either TRUE or FALSE), which indicates whether the output should be ordered or not.

Display data frame
```{r}
study_data
```

R Code to test the claim that energy levels are the same for students in the population regardless of energy drink consumption level.

- $H_0$: no change in energy level regardless of energy drink consumption level
- $H_1$: Not $H_0$

- $\alpha = 5$% (0.05)
```{r}
anova2 <-aov(energy_level ~ consumption_level, data=study_data)

summary(anova2)
```

Alternative R code is to run the following:
```{r}
oneway.test(energy_level ~ consumption_level, data=study_data, var.equal=TRUE)
```

Note that the assumptions for one-way ANOVA include the following:

1. Variances are are equal (or homogeneous)
2. Samples/observations are independent
3. dependent variable should be measured on at least an interval scale


# Case 2: ANOVA for dependent samples - i.e. more than two matched samples

This type of ANOVA is also called repeated measures or randomized blocks or within-subjects


## Example 1

Repeated measures ANOVA in R

```{r}
# Suppose we have the following people sign up for a weight-loss program
person <-c("Al", "Bill","Charlie","Dan", "Ed", "Fred","Gary", "Harry", "Irv", "Jon")

# Suppose their respective weight measurements before entering the program are:
before <-c(198,201,210,185,204,156,167,197,220,186)

# Suppose their respective weight measurements after 1 month in the program are:
one_month <-c(194,203,200,183,200,153,166,197,215,184)

# Suppose their respective weight measurement after 2 months in the program are:
two_month <-c(191,200,192,180,195,150,167,195,209,179)

# Suppose their respective weight measurement after 3 months in the program are:
three_month <-c(188,196,188,178,191,145,166,192,205,175)

# Create a data frame to hold the data
initweight_df <-data.frame(person, before, one_month, two_month, three_month)

# View the final data frame containing weight measures for the participants:
initweight_df
```

Note that the data is in wide format.
We need to reshape the data into long format.
We use the `melt()` function to convert the data frame from wide format to long format.
Make sure to install and load the "reshape" package.
```{r more packages, warning = FALSE, message = FALSE}
#install.packages("reshape")
library("reshape")
```
```{r}
finalweight_df <-melt(initweight_df, id="person")

# Label the new columns appropriately
colnames(finalweight_df) = c("person", "time","weight")

# Display the new data frame
finalweight_df
```
Conduct a repeated measures ANOVA at 5% significance level

We can use ANOVA to check the following two cases based on the data frame `finalweight_df`

- **Case A:** Check if weight across time is the same or not in the population regardless of person

- **Case B:** Check if weight across time is the same or not in the population per person/accounting for person

## Case A. `aov(weight ~ time, data = finalweight_df)`

- $H_0$: weight before and across time (i.e. one_month, etc) are equal/ no change from before to other periods
- $H_1$: Not $H_0$

- $\alpha =5$% (0.05)
```{r}
anova1 <-aov(weight ~ time, data=finalweight_df)

anova1

summary(anova1)
```

***Finding/result:***

p-value (0.634) > alpha (0.05). 

We fail to reject the null hypothesis $H_0$

***Conclusion:***

There is evidence that weight before and across time (i.e. one_month, etc) are equal/ no change

## Case B. `aov(weight ~ time + Error(person/time),data = finalweight_df)`

- $H_0$: weight before study and across time (i.e. one_month, etc) after study are equal/no change from before to other periods per person
- $H_1$: Not $H_0$

- $\alpha =5$% (0.05)
```{r}
anova2 <-aov(weight ~ time +Error (person/time), data=finalweight_df)
```

The code above allows us do a "within-subjects" ANOVA analysis

Display ANOVA results
```{r}
summary(anova2)

summary(anova1)
```

***Finding/result:***

p-value (7.3e-08) < alpha (0.05). 

We reject the null hypothesis $H_0$

***Conclusion:***

There is evidence in favor of the alternative 

i.e. there is evidence that weight before and across time (i.e. one month, etc) are not equal per person across time



# %Session Info%
```{r session info}
sessionInfo()
```