---
title: "Two Sample Means Testing"
format:
    html:
        code-fold: false
execute:
    output: true
    warning: false
toc: true
---
```{r libraries, warning=FALSE,message=FALSE}
#Load Tidyverse
library(tidyverse)

#load dplyr
library(dplyr)

#load ggplot2
library(ggplot2)
```

Today we discuss two-sample hypothesis testing for differences in population means. 
In a variety of fields, the need often arises to compare the behavior of one population with the behavior of another population.
To do this, we draw two samples (one from each population). 
Hence the name "Two Sample Tests".

Sometimes the samples are independent, and sometimes they are matched in some way. Each sample comes from a separate population

The objective is mostly to determine whether these populations are different from each other. Usually this involves tests of hypotheses about population means.

We can also test hypotheses about population variances.

In testing hypotheses with two samples (from two different populations).
The null hypothesis ($H_0$) here is that the two populations are similar

i.e. the average behavior of (a variable) in the two populations are similar.

i.e. any differences you see between the two samples are due to chance and not real.

The alternative hypothesis ($H_1$) is that the average behavior of (a variable) in the two populations is NOT similar/the same.

We can have two-sided hypothesis testing for means:

**null**        $H_0$: mean 1 = mean 2 or mean 1 - mean 2 = 0
**alternative** $H_1$: mean 1 $\neq$ mean2  or mean 1 - mean 2 $\neq$ 0

We can also have one-sided or one-tailed tests here, where the

**null** $H_0$: mean 1 - mean 2 = 0 
and the **alternative**,
      $H_1$: mean 1 - mean 2 > 0 

**null** $H_0$: mean 1 - mean 2 = 0 
and the **alternative**,
      $H_1$: mean 1 - mean 2 < 0


The zero in these hypotheses is just the typical case.
It is possible to test for any value - just substitute that value for zero e.g.

where the **null** $H_0$: mean 1- mean 2 = 5 
and the **alternative**,
               $H_1$: mean 1 - mean 2 $\neq$ to 5




**To carry out the test follow these steps:**

Step 1: Set significance level (select your alpha)

Step 2: Write out the null and alternative hypotheses

Step 3: Calculate the mean and standard deviation of each sample

Step 4: Subtract one mean from the other

Step 5: Use a formula to convert the result of Step 4 into a test statistic

Step 6: Obtain the p-value corresponding to the statistic

Step 7: Compare the p-value from Step 5 to your chosen alpha

Step 8: If p-value < = alpha reject null hypothesis otherwise fail to reject the null hypothesis

Step 9: Write your results and conclusion about the test

# Two sample tests for means in R

There are two different t-tests:

1. ***Independent-means t-test***: also called the *independent-measures* or *independent-samples t-test* (assume equal or unequal variances)
2. ***Dependent-means t-test***: also called *matched-pairs* or *paired-samples t-test*

# INDEPENDENT MEANS T-TEST: Unknown population variances but are assumed equal
## Example 1: Working with samples from two populations. 
Assume equal variances

Using 2 samples from 2 populations.
The variances of the population variables are not known but are assumed equal

When you do not know a population variance, you use the sample variance to estimate it.

When you have two samples, you average the two sample variances to arrive at the estimate (this is called pooling)

Assume we have the following samples from populations we do not know variances for.
Assume the variances are equal.

The data are from production times for two machines (machine1 and machine2).
The company is trying to determine if the average production time for both machines is the same (i.e. difference in population means for production time = 0)
```{r}
machine1 <-c(24.58,22.09,23.70, 18.89, 22.02,28.71,24.44,20.91,23.83,20.83)

machine2 <-c(21.61,19.06,20.72,15.77,19,25.88,21.48,17.85,20.86,17.77)
```

To test the following hypotheses at 5% significance level($\alpha$):

- $H_0$ (null): true difference in means = 0
- $H_1$ (alternative): true difference in means <> 0 (not equal to zero)

**R Code to test $H_0$ versus $H_1$ above:**
```{r}
t.test(machine1, machine2, var.equal=TRUE, alternative="two.sided", mu=0)
```
***Finding/Result:***

p-value (0.02528) is less than 5% significance setting (i.e. p-value $< \alpha$) so we reject the null hypothesis.

***Conclusion:***

Since we reject the null, there is evidence that the true difference in means (i.e. the difference in the population variable means) is not equal to zero.
This means that the two populations are different or behave differently.

## Example 2: Working with a data frame. Assume variances are equal

We can work with a data frame that contains the two machine data as follows:

Create data frame in long format. First create a vector for the 20 production times - machine1's times first then machine2's production times.
```{r}
prod.time <-c(machine1, machine2)

#create a vector of the two machine names
machine <- c("machine1", "machine2")

#Turn vector above into a vector of 10 repetitions of machine1 and machine2
machine <-rep(machine, times=c(10,10))

#Create data frame
DF <-data.frame(machine, prod.time)

#display DF
DF
```
```{r, eval = FALSE}
#or use the view function
view(DF)
```

We can test the following at $\alpha = 5$%

- $H_0$: Mean production time across the two machines are similar/equal
- $H_1$: Mean production time across the two machines are not similar/equal


We then conduct the t-test as follows:
```{r}
with(DF, t.test(prod.time ~ machine, var.equal=TRUE,
                alternative="two.sided", mu=0))
```

***Finding/result:***

the p-value (0.02) is < alpha (0.05), so we reject the null hypothesis ($H_0$)

***Conclusion:***

Since we reject the null hypothesis, there is evidence in favor of the alternative (i.e. there is evidence that the mean production time
 for the two machines are not similar/equal)

visualizing your findings/results in a Boxplot:
```{r}
ggplot(DF, aes(x=machine, y=prod.time))+
  stat_boxplot(geom="errorbar", width=.5)+
  geom_boxplot()
```

## Example 3

The data are from production times for two performance monitoring devices (fitbit1 and fitbit2). Suppose you are trying to determine if the devices are very similar in the population (i.e. difference in population means for heartbeat measurements = 0).
```{r}
fitbit1 <-c(65,120,45, 46, 47,49,51,70,44,69, 85,81, 80)

fitbit2 <-c(61,109,45,55,66,85, 62,74,76,66,85,102,67)
```

To test the following hypotheses at 1% significance level(alpha):

- $H_0$ (null): true difference in means = 0
- $H_1$ (alternative): true difference in means <> 0 (not equal to zero)

**R Code to test $H_0$ versus $H_1$ above:**
```{r}
t.test(fitbit1, fitbit2, var.equal=TRUE, alternative="two.sided", mu=0)
```

***Finding/Result:***

p-value (0.3382) > alpha (0.01). We fail to reject the null hypothesis

***Conclusion***

Since we fail to reject the null hypothesis, we conclude that there is evidence in favor of the alternative hypothesis.

## Example 4

The data are from production times for two performance monitoring devices (fitbit1 and fitbit2).
Suppose you are trying to determine if the devices are very similar in the population (i.e. difference in population means for heartbeat measurements = 5).
```{r}
fitbit1 <-c(65,120,45, 46, 47,49,51,70,44,69, 85,81, 80)

fitbit2 <-c(61,109,45,55,66,85, 62,74,76,66,85,102,67)
```

To test the following hypotheses at 1% significance level(alpha):

- $H_0$ (null): true difference in means = 5
- $H_1$ (alternative): true difference in means <> 5 (not equal to 5)

**R Code to test $H_0$ versus $H_1$ above:**
```{r}
t.test(fitbit1, fitbit2, var.equal=TRUE, alternative="two.sided", mu=5)
```

***Finding/Result:***

p-value (0.1214) > alpha (0.01). We fail to reject the null hypothesis

***Conclusion:***

Since we fail to reject the null hypothesis, there is evidence in favor of the alternative

## Example 6: One-sided test/one-tailed

Suppose you have two samples from two groups of students that took the same test and you want to test which population did better on the test on average.
Assume the population variances are equal for the test score variable.

Conduct a significance test at 10% that first set of test takers (`test_t1`) are better test takers than the second set of test takers (`test_t2`)
```{r}
test_t1 <-c(87,85,99,75,92,43,56)
test_t2 <-c(98,97,64,56,98,100,47,66)
```

- $H_0$: population mean for `test_t1` = population mean for `test_t2`
- $H_1$: population mean for `test_t1` > population mean for `test_t2`
- $\alpha = 10$% = 0.1
```{r}
t.test(test_t1, test_t2, var.equal=TRUE, alternative="greater", mu=0)
```

***Findings/results?***


***Conclusion?***

# INDEPENDENT MEANS T-TEST: Unknown population variances but are NOT assumed equal

## Example 1: Assume unequal population variable variances 

The data are from production times for two machines (machine1 and machine2). 
The company is trying to determine if the average production time for both machines is the same (i.e. difference in population means for production time = 0)
```{r}
machine1 <-c(24.58,22.09,23.70, 18.89, 22.02,28.71,24.44,20.91,23.83,20.83)

machine2 <-c(21.61,19.06,20.72,15.77,19,25.88,21.48,17.85,20.86,17.77)
```

To test the following hypotheses at 5% significance level(alpha):

- $H_0$ (null): true difference in means = 0
- $H_1$ (alternative): true difference in means <> 0 (not equal to zero)

**R Code to test $H_0$ versus $H_1$ above:**
```{r}
t.test(machine1, machine2, var.equal=FALSE, alternative="two.sided", mu=0)
```
***Finding/Result:***

p-value (0.02528) is less than 5% significance setting (i.e. p-value < alpha) so we reject the null hypothesis.

***Conclusion:***

Since we reject the null, there is evidence that the true difference in means (i.e. the difference in the population variable means) is not equal to zero.
This means that the two populations are different or behave differently.


Given that we reject the null for both significance tests (when we assume equal and unequal variances), we can conclude that the production times for both machines are different at the population level (i.e. their average behavior is different).


## Example 2: Difference in means test with unknown population variances that are assumed equal

Suppose the Boulder Valley School District (BVSD) devised a questionnaire to measure the level of satisfaction for graduates of two competing  state schools. Past history indicates that the variance in the metric for measuring satisfaction for both schools is equal.

The statistic is randomly administered to 11 schools from the first school (School1) and 15 students from the second school (School2) as follows:
```{r}
school1 <-c(69,75,76,80,81,82,86,89,91,92,97)

school2 <-c(59,62,66,70,70,75,75,77,78,79,81,84,84,86,94)
```

### Question 1:
Conduct a test of significance at 10% to verify if the difference in mean population satisfaction levels is greater than zero. Assume the variances in satisfaction levels in the population are not known and are not known to be equal.

***Question 1 Answer:***

This is a one-sided test

- $H_0$: Mean population satisfaction levels for school 1 is equal to mean population satisfaction levels for school 2 (implying that difference mean population satisfaction levels is less than or equal to zero).
- $H_1$: Mean population satisfaction levels for school 1 is greater than population satisfaction levels for school 2 (implying that difference mean population satisfaction levels is greater than zero).
- $\alpha = 0.1$

**R code for the test:**
```{r}
t.test(school1, school2, var.equal=FALSE, alternative="greater", mu=0)
```

***Findings:***

p-value < 0.05. 

So we reject the null hypothesis

***Conclusion***

Since we reject the null hypothesis, there is evidence that the Mean population satisfaction levels for school 1 is greater than population satisfaction levels for school 2 (implying that difference mean population satisfaction levels is greater than zero).

### Question 2
Conduct a test of significance at 5% to verify if the difference in mean population satisfaction levels is zero or not equal to zero.
Assume the variances in satisfaction levels in the population are **NOT** known to be equal.

***Question 2 Answer:***

This is a two-sided test

- $H_0$: Mean population satisfaction levels are equal (implying that difference mean population satisfaction levels is equal to zero)

- $H_1$: Mean population satisfaction levels are not equal (implying that difference mean population satisfaction levels is not equal to zero)

- $\alpha = 0.05$

**R code for the test:**
```{r}
t.test(school1, school2, var.equal=FALSE, alternative="two.sided", mu=0)
```

***Finding/Result:***

p-value < 0.05

So we reject the null hypothesis

***Conclusion:***

Since we reject the null hypothesis, there is evidence that Mean population satisfaction levels are not equal (implying that difference mean population satisfaction levels is not equal to zero)

i.e. there is evidence in favor of the alternative

### Question 3
Conduct a test of significance at 1% to verify if the difference in mean population satisfaction levels is zero or not equal to zero.
In this question, assume historical variances are not known to be equal.

***Question 3 Answer:***

This is a two-sided test

- $H_0$: Mean population satisfaction levels are equal (implying that difference mean population satisfaction levels is equal to zero)

- $H_1$: Mean population satisfaction levels are not equal (implying that difference mean population satisfaction levels is not equal to zero)

- $\alpha = 0.01$

**R code for the test:**
```{r}
t.test(school1, school2, var.equal=FALSE, alternative="two.sided", mu=0)
```

***Finding/Result:***

p-value < alpha(0.05)

So we reject the null hypothesis

***Conclusion:***

Since we reject the null hypothesis, there is evidence that Mean population satisfaction levels are not equal (implying that difference mean population satisfaction levels is not equal to zero)

### Question 4
Conduct a test of significance at 1% to verify if the difference in mean population satisfaction levels is equal to or not equal to 70.
In this question, assume historical variances are not known to be unequal.

***Question 4 Answer:***

This is a two-sided test

- $H_0$: Mean population satisfaction levels are equal to 70 (implying that difference mean population satisfaction levels is equal to 70)

- $H_1$: Mean population satisfaction levels is not equal to 70 (implying that difference mean population satisfaction levels is not equal to 70)

- $\alpha = 0.01$

**R code for the test:**
```{r}
t.test(school1, school2, var.equal=FALSE, alternative="two.sided", mu=70)
```

***Finding/Result:***

p-value (6.156e-15) < alpha (0.05)

So we reject the null hypothesis

***Conclusion:***

Since we reject the null hypothesis, there is evidence that the mean population satisfaction levels is not equal to 70 (implying that difference mean population satisfaction levels is not equal to 70)

i.e. there is evidence in favor of the alternative

# DEPENDENT MEANS T-TEST/PAIRED SAMPLES T-TEST

Independent means t-test uses samples that are not dependent on each other 
(i.e. choosing a member of one sample has no bearing on choosing a member for the other sample)

Dependent means t-tests use samples that are dependent or matched in one way or another 

e.g. in a before-and-after study, members of one sample (before) are the same members that are selected for the other sample(after).

For example, a weight-loss before/after study would draw a sample from a population but effectively reuse the members of one sample in the other sample.

## Example 1

A pharmaceutical company is running a weight-loss program.
The belief of a skeptical newspaper is that the program does not deliver any benefits to members of a community.

To prove their point, the newspaper hires a research student to gather the following weight data from participants in the program:
```{r}
before <-c(198,201,210,185,204,156,167,197,220,186)
after <-c(194,203,200,183,200,153,166,197,215,184)
```

The newspaper asks the student to test the following hypotheses:

- $H_0$: mean population weight before is equal to mean population weight after participating in program
- $H_1$: mean population weight before is not equal to mean population weight after participating in program

- $\alpha = 5$% (0.05)
```{r}
t.test(before, after, alternative="two.sided", paired=TRUE)
```

***Finding/Result:***

p-value < alpha, implying that we reject the null hypothesis

***Conclusion:*** 

Since we reject the null hypothesis, we conclude that there is evidence in favor of the alternative hypothesis.

i.e. there is evidence that the mean population weight before participating in the weight-loss program is not equal to the mean population weight after participating in the weight-loss program.

We can also do a one-sided, paired sample t-test using the same data as follows:

- $H_0$: mean population weight before is less than or equal to mean population weight after participating in program
- $H_1$: mean population weight before is greater than mean population weight after participating in program

- $\alpha = 5$% (0.05)
```{r}
t.test(before, after, alternative="greater", paired=TRUE)
```

***Finding/Result:***

p-value < alpha, implying that we reject the null hypothesis

***Conclusion:***

There is evidence that the mean population weight before participating in the weight-loss program is greater than the mean population weight after  participating in the weight-loss program


# %Session Info%
```{r session info}
sessionInfo()
```