---
title: "Intro to R-Markdown Solutions"
author: "Dr. Osita Onyejekwe"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 2: Data Wrangling and Tidy Data

During today's recitation we will focus on data wrangling, including importing data, changing data types, subsetting variables and observations, and calculating basic functions. 

## Question 1: Importing Data
Download the California foreclosure auctions dataset and save it to a folder on your desktop. Then load the data into R and save it as a data frame.

```{r }
library(tidyverse)
# load data 
df <- read.csv("../California Foreclosure Auctions_updated.csv")
```

## Question 2: Check out the Data
What is the observational unit of this data? Is the data in tidy format? What needs to be cleaned up about this data?
```{r }
# make a copy of the data so we don't overwrite the original when we make changes
df_clean <- df

# make zip code a factor instead of an integer
df_clean$Zip <- as.factor(df$Zip)
```

```{r}
# convert transfer amount to an integer instead of a character
library(readr)
df_clean$TransferAmount <- parse_number(df$TransferAmount)

# check for missing values in the varaibles of interest
which(df$RadarID=="")

```

```{r}
which(is.na(df$Zip))

```

```{r}
# drop observations with missing data
df_clean <- df_clean %>% filter(df$RadarID != "")
```

This dataset is pretty clean: the import successfully identified column names that are easy to use, the variables corresponding to demographic information were imported as characters (`RadarID`, `County`, `Address` etc…) and most of the numeric variables were imported as numeric variables (`Units`, `SqFt`, `YearBuilt` etc…).

However, it looks like Zip code was imported as an integer. This is troublesome because it will allow mathematical operations on zip codes which doesn’t make sense (for example multiplying a zip code times 2 has no meaning) so we want to make that a factor variable. Also, the `TransferAmount` variable was imported as a character so we will need to change this to a numeric.

Lastly, there are some missing values in the last few rows. In the code above, the `which()` function returns the row numbers that have missing values in the columns `RadarID` or `Zip`.


## Question 3: Subset Rows and Columns
Create a new data frame that contains only the auctions that are "TrusteesDeed-3rd" type, and only the variables required to uniquely identify each observation, along with "TransferAmount". (Bonus: make a scatter plot of the estimated value vs the final sale price and color code it by County - make some interpretations)

```{r }
# how to do this using tidyverse
library(tidyverse)
df_small <- df_clean %>%
            filter(TransferType == "TrusteesDeed-3rd") %>%
            select(RadarID, County, TransferAmount)

# another way to do this using base r
df_small <- df_clean[df_clean$TransferType == "TrusteesDeed-3rd", c("RadarID","County","TransferAmount")]

```

## Question 4: Tidy the Data
If we wish to treat the County as the observational unit, then we must tidy the data. We can do this by averaging across all observations within the new observational unit. Creat a new data.frame that contains the County variable along with average transfer price.

```{r}
# how to do this using tidyverse
df_tidy <- df_clean %>%
           group_by(County) %>% 
           summarise(AvgPrice = mean(TransferAmount, na.rm=TRUE))
head(df_tidy)
```

```{r}
# another way to do this using base r
df_tidy <- aggregate(TransferAmount ~ County, data = df_clean, FUN = mean)
head(df_tidy)
```


## Question 5: Calculate By Groups
What is the average sale price for TrusteesDeed-3rd transfers for each county? Which county is the most expensive?
```{r }
df_small %>% group_by(County) %>% summarise(AvgPrice = mean(TransferAmount, na.rm=TRUE))

```

```{r}
# another way using base r
aggregate(TransferAmount ~ County, data = df_small, FUN = mean)
```


## Final Questions:

1) What was the main point of today's content?

2) What is the muddiest point of the class so far?
